Mitochondrial genomes of Pleistocene megafauna retrieved from recent sediment layers of two Siberian lakes

Version 1

Abstract

Abstract

Ancient environmental DNA (aeDNA) from lake sediments has yielded remarkable insights for the reconstruction of past ecosystems, including suggestions of late survival of extinct species. However, translocation and lateral inflow of DNA in sediments can potentially distort the stratigraphic signal of the DNA. Using three different approaches on two short lake sediment cores of the Yamal peninsula, West Siberia, with ages spanning only the past hundreds of years, we detect DNA and assembled mitochondrial genomes of multiple mammoth and woolly rhinoceros individuals—both species that have been extinct for thousands of years on the mainland. The occurrence of clearly identifiable aeDNA of extinct Pleistocene megafauna (e.g., > 400K reads in one core) throughout these two short subsurface cores, along with specificities of sedimentology and dating, confirm that processes acting on regional scales, such as extensive permafrost thawing, can influence the aeDNA record and should be accounted for in aeDNA paleoecology.

Introduction

Sedimentary deposits constitute highly valuable archives of past ecosystem changes as they contain dateable layers with organismic remains including ancient DNA (aDNA). Such remains are typically assumed to represent the ecosystem of the time around which the respective stratum was deposited. aDNA from sediments has yielded remarkable insights regarding paleoecology, phylogeography, and extirpation and extinction events of keystone taxa such as mammoths (Haile et al., 2009; Boessenkool et al., 2012; Graham et al., 2016; Murchie, Monteath, et al., 2021). Based on such ancient environmental DNA (aeDNA), a recent study proposed that the wooly mammoth (Mammuthus primigenius) may have survived in Eurasia for much longer than previously assumed, as the authors retrieved mammoth DNA sequences in sediment layers that were approximately 4.6—7 thousand years (kyr) younger than the most recent mammoth fossils (Wang et al., 2021); however, in response to this interpretation, Miller and Simpson (2022) opined that these results may be more likely due to taphonomic processes leading to release of aDNA from the remains of long-dead organisms from permafrost, where it is well preserved.

Conclusions derived from aDNA isolated from sediment cores rely on the stratified nature of the remains in question and dating of the respective layer by radiometric methods. However, in theory, various physical and geochemical processes such as translocation of DNA through sediment strata (Haile et al., 2007), re-deposition of older sediment carrying DNA of extinct organisms (Arnold et al., 2011), and preservation bias (Boere et al., 2011) can distort the biological signal of aDNA and thus bias the accuracy of allocation of taxa to specific time periods (Armbrecht et al., 2019). For lake sediments, deposited under aquatic conditions, studies have suggested that leaching is not a concern (Parducci et al., 2017), but it has been observed in soils and cave sediments (Haile et al., 2007). The question of obtaining last appearance dates of extinct taxa using aDNA in dated sediment layers (Haile et al., 2009) is under discussion, but the surprisingly young records published to date still date to multiple thousands of years before present and thus lie within a timeframe of possible late survival.

Results and Discussion

In 2019, we retrieved short subsurface sediment cores from two Arctic thermokarst lakes (LK-001 and LK-007, located approximately 5 km apart, over massive permafrost; Table 1; Dvornikov et al., 2016) on the Yamal peninsula, Siberia, to extract DNA and assess changes in mammal abundances in the Arctic over the past decades and centuries. From lake LK-001, we collected a secondary core which was sliced in the field at 1-cm steps for Pb210 radiometric dating, which indicated that the sediments at the top of this core were deposited recently, and that the core spanned the past few centuries (Table S5). The cores for DNA extraction were closed in the field immediately after retrieval and were then transported to the dedicated aDNA laboratories of the University of Konstanz, Germany. In this lab facility, established in 2020, no other samples from the Arctic or from any large mammals had been processed previously. From core LK-001, we isolated DNA and produced genomic double-stranded libraries from 23 samples according to standard procedures. The core was opened and all subsequent steps until index PCR setup were carried out under customary aDNA laboratory conditions. In particular, the core opening facilities and the lab are located in buildings separated from the downstream molecular genetic analyses, the ventilation of the aDNA lab is based on a HEPA filter system and positive air pressure, and the lab is subjected to nightly UV radiation. Work in the lab is conducted under strict aDNA precautions, adhering to established aDNA protocols (Fulton and Shapiro, 2019). We enriched the libraries for mammalian DNA using a custom RNA bait panel produced from complete mitogenome sequences of 17 mammal species that currently or previously occurred in the Arctic (adapted from Murchie, Kuch, et al., 2021). The enriched libraries were sequenced, and we mapped the sequences against a database of 73 mammal mitogenomes, followed by BLASTn alignment against the complete NCBI nt database. We thus retrieved mitogenomic sequences of mammals that were expected during the age range covered by the core, e.g., reindeer (Rangifer tarandus), Arctic lemming (Dicrostonyx torquatus), and hare (Lepus); however, throughout the entire core, there were abundant sequences of two species that have been extinct for several thousand years, i.e., mammoth (Mammuthus primigenius) and woolly rhinoceros (Coelodonta antiquitatis). Twenty-two of the 23 LK-001 libraries produced > 1,000 reads, each, assigned to Mammuthus, with read counts ranging from 2,852 to 72,919 (mean 21,140 ± 17,296). Negative controls (extraction and library blanks) did not produce any reads assigned to mammals. In the sample with the highest M. primigenius read counts (31.5 cm depth, dated to 81 years), the coverage of the reference mitogenome (NCBI accession NC_007596.2) was 95.3%, (434 (± 213)-fold). Across all samples, 465,080 reads assigned to mammoth were produced, with 98.3% coverage (2,762-± 1,176-fold). Read lengths ranged from 28 to 289 bp (mean 100 ± 44 bp; Fig. 1). The number of wooly mammoth reads decreased from lower samples towards the top of the core (Fig. 1), and mapping suggested that the sequences throughout the core originated from multiple individuals. Signatures of post-mortem DNA decay were comparably minor (Fig. 1), with reference to an M. primigenius genome downloaded from NCBI (accession NC_007596.2). The proportions of bases did not suggest a substantial deviation from those in the reference genomes or in the closest extant relative of Mammuthus, the Asian elephant (Elephas maximus). Twelve of the 23 libraries produced > 100 reads, each, assigned to wooly rhinoceros, with a total of 2,737 reads and a cumulative coverage of 44% (assembled to NC_012681.1 2).

aDNA of Mammuthus in recent lake sediments. (A) Read counts assigned to Mammuthus (square-root-transformed proportion of the respective number of raw reads per library) after hybridization capture enrichment of ancient DNA of core LK-001 (shown are results of 22 libraries; one library was excluded as it did not produce any reads assigned to mammals). Sample depths are indicated on the horizontal axis (in cm); the solid line indicates the general trend. (B) Fragment length distribution and (C) damage patterns across the 22 libraries.

As these results were entirely incongruent with the temporal occurrence of these species, we employed several methods to confirm their validity, i.e., conventional PCR and Sanger sequencing of a mammoth COI fragment, mammal metabarcoding, and droplet digital PCR (ddPCR) of a mammoth cytochrome b fragment. We performed these analyses on core LK-001 and, to evaluate whether this is a locally isolated phenomenon, on a second short core, LK-007, from a nearby lake, which had been opened and processed under the same conditions indicated above to prevent contamination. Additionally, we screened the core LK-001 for plant macrofossils, of which three were sent for 14C AMS dating.

Conventional PCR and Sanger sequencing confirmed amplification of a COI fragment of M. primigenius. Mammal metabarcoding produced mammoth sequences (74 bp) in 13 samples of core LK-001 and 9 samples of core LK-007. In the LK-001 core, the highest M. primigenius read count occurred at 31.5 cm (2,992 reads), and in the LK-007 core at 26 cm M. primigenius (3,580 reads). ddPCR produced M. primigenius sequences in 14 samples of each core. Metabarcoding and ddPCR patterns across the cores were similar, although not completely congruent, as ddPCR appeared to be more sensitive (Fig. S1). The dates retrieved by radiocarbon dating were not congruent with the initial age inference suggested by Pb210. While the lowest and topmost sample (with ages of 1,547 ± 228 and 339 ± 79 uncal. yrs BP respectively) suggest relatively young ages and agree in their temporal succession, the middle sample, at 36,5 cm, yielded a radiocarbon age of 8,677 +/-132 yrs.

The mammoth was abundant throughout most of Eurasia during the Pleistocene, but populations declined at the end of the Pleistocene, with the species going extinct in the mid-Holocene (Nogués-Bravo et al., 2008). The youngest fossils from mainland Siberia have been dated to 9,650 years (Stuart et al., 2002). The wooly rhinoceros was a cold-adapted megaherbivore, which was abundant from western Europe to north-east Siberia during the Middle to Late Pleistocene (Kahlke and Lacombat, 2008). This species was predominantly grazing, probably resorting to browsing only due to seasonal vegetation restrictions (Kahlke and Lacombat, 2008; Rey-Iglesia et al., 2021; Stefaniak et al., 2021). The reasons underlying the extinction of this species at ca 14 ka BP are not entirely clear, but it is largely attributed to sudden climate warming and subsequent habitat unsuitability due to vegetation changes (Stuart and Lister, 2012; Lord et al., 2020; Wang et al., 2021) in the Bølling-Allerød interstadial warming.

The present data, which implies frequent and abundant Pleistocene megafaunal DNA throughout a sediment core deposited in the lake over the past centuries suggests that physical processes, rather than presence of live organisms, are responsible for the recovery of this DNA. While not in itself fully conclusive, our data suggests the source of the DNA of Pleistocene mammals from older permafrost deposits, either from a carcass or from sedimentary materials carrying the DNA. The numbers of mammoth sequences increased with depth downcore, with comparably low abundances over the most recent 11 cm of the core (aged < 30 years), pointing to a decrease of input of this DNA through time. The apparently rather limited extent of aDNA damage in the mammoth sequences suggests that the source of this DNA has been preserved exceptionally well, which also suggests an origin from permafrost, and the specific dynamics of thawing and redeposition of material in the study area offer an explanation.

Here, the active thermokarst likely began during the climatic optimum of the Holocene (9000–3000 BC; (Savvichev et al., 2021). The LK-001 and LK-007 lake basins are interbedded mainly into the IVth marine plain, formed in the Kazantsevskaya, Marine Isotopic Stage 5. The permafrost of these lake catchments was formed no earlier than 70—60 kyr BP after the Kazantsevskaya transgression in more sub-aerial conditions and with mean annual temperatures 6 to 7 °C lower than modern temperatures (Baulin et al., 1981). As this area was under coastal-marine conditions for a long period, these lake basins may be paleo-marine remnants, or they were formed later as a result of thermokarst over the segregated or tabular (massive) ground ice during the Holocene climatic optimum (Kachurin, 1961). The area is also subject to abrupt permafrost thaw (thermo-denudation), resulting, for example, in the formation of retrogressive-thaw slumps (thermocirques) and the transport of a large amounts of thawed terrestrial material into the lake water (Dvornikov et al., 2018). Such abrupt permafrost thaw processes normally appear adjacent to lakes and can form specific geomorphological elements,. i.e., thermo-terraces (Kizyakov et al., 2006) within lake catchments and lakes; they are normally polycyclic processes appearing due to the extension of a seasonally thawed layer (active layer) up to the top of massive ice and more favorable thermal conditions within the existing forms (Leibman and Kizyakov, 2007; Kokelj et al., 2009). Traces of past thermo-denudation can be observed within both lake catchments. In catchments of five neighboring lakes, large retrogressive thaw slumps appeared in recent years (2012-2013) accompanied by the thaw and lateral transport of modern and late Pleistocene deposits into lakes. Additionally or alternatively, thermo-erosion of upper geomorphological levels and transport via stream networks could transport ancient material into the modern lacustrine sediments. However, the two studied lakes are headwater lakes (with outlet, no apparent inlet) and this option can only be considered in terms of small thermo-erosional valleys within the catchments.

An alternative mechanism for the redistribution of Late Pleistocene material in the sediments is related to subcap methane emission (bubbling) from degrading permafrost beneath the lake bottom. In-lake bubbling can be observed in a circum-Arctic scale: in North-East Siberia, Alaska and Canada. This is common especially in lakes with a depth exceeding two meters, which do not freeze entirely up to the bottom in winter, leading to the formation of a talik. The expansion of the talik may further trigger subcap methane emission, which can reach 40-70 kg yr-1 of pure (94-100%) methane in neighboring lakes (Kazantsev et al., 2020). The constant methane seepage doesn’t allow ice to be formed in winter (whereas the normal winter ice thickness is approximately 1.5 m) and can potentially disturb the stratigraphy of lake sediments. Additionally, dramatic emissions of methane can form craters in terrestrial and lacustrine environments (Dvornikov et al., 2019). In this case Late Pleistocene sediments will well be re-distributed within the water-body and the entire stratigraphy will be mixed.

In the case reported here, the simultaneous finding of a Pb210 chronology indicating recent sediment deposition and of plant macrofossils that dated to > 8000 yrs BP in a sample from 36.5 cm, suggest lateral input of ancient material, including the mammalian DNA, putatively related to permafrost thawing processes. Numerous studies on aDNA discussed possible leaching through sedimentary strata of the DNA itself, yet it was typically considered not an issue as most of these studies were conducted under stable permafrost or similar conditions (eg, Hansen et al., 2006; Haile et al., 2007, but see Andersen et al. 2012). Permafrost thawing and redeposition of material adds a new dimension to this problem of temporal interpretation. The fact that we retrieved mammoth sequences from both cores of two lakes located approximately 5 km apart suggests that this is not an isolated phenomenon but occurs on a regional or even larger scale. Given the wide spread of abrupt permafrost thaw processes in the Arctic plains (West Siberia, Taimyr, Chukotka, Alaska, Canadian Arctic (Kizyakov et al., 2006 and references therein), the phenomena of disturbed stratigraphy of lacustrine sediments can potentially be observed at a pan-Arctic scale. While this indicates that temporal interpretation of sedimentary aDNA records should be exercised with caution, our study also demonstrates that a careful evaluation of available information on the site and ecosystem in conjunction with the use of independent dating techniques can uncover incongruencies. This is more difficult in older time periods, where artefactual stratigraphies caused by equivalent processes acting for a limited time will not be detected as easily as in our case of long extinct species. The same applies to extant taxa or those which have undergone extinction or extirpation more recently, the presence of which cannot as easily be excluded as in the current example. However, we suggest that the inclusion of robust dating techniques and knowledge of local geophysical processes can provide good arguments to evaluate the reliability of aeDNA records.

Materials and Methods

Field sites, DNA isolation and hybridization capture enrichment

In 2019, sediment cores (6 cm diameter; Table 1) were retrieved from two lakes (LK-001 and LK-007, respectively, Table 1) on the Yamal peninsula, Siberia, using a UWITEC piston corer (UWITEC, Mondsee, Austria). The lakes were located approximately 5 km apart. The cores were transported to the aDNA laboratories of the University of Konstanz, Germany; from lake LK-001, a secondary core was taken which was sliced in the field at 1-cm steps for radiometric dating from 0 to 39.5 cm depth, performed at the Environmental Radioactivity Research Centre, University of Liverpool (supplementary material). Additional 14C dating of three specimens of plant remains, extracted at 36.5 cm, 51 cm and 74 cm, was performed (supplementary material).

Sedimentary DNA was isolated from 23 samples of core LK-001 and from 16 samples of core LK-007 using commercially available kits with modified protocols (supplementary material). The extracts of core LK-001 were subjected to library preparation for capture enrichment. Enrichment probes were designed from mitogenomes of 17 herbivorous mammal species that currently or previously occurred in the Arctic (Table S1) and few lichen sequences (Table S2). Genomic libraries were produced according to (Li et al., 2013) with some modifications (Seeber et al., 2019) (see supplementary material). Filtered reads were mapped to mammalian mitogenomes, followed by BLASTn alignment against the complete NCBI nucleotide database and subsequent metagenomic analyses using MEGAN (Huson et al., 2016). Reads assigned to Mammuthus were mapped to a complete M. primigenius reference mitogenome (NCBI accession NC_007596.2); reads assigned to Coelodonta antiquitatis were mapped to the NCBI reference mitogenome NC_012681.1 2.

Conventional PCR, mammal metabarcoding, and ddPCR

Based on the enriched fragments with the highest coverage, PCR primers specific to M. primigenius were designed using Geneious Prime 2022.1.1 (Kearse et al., 2012), i.e., mamm801 (5‘-CCCATGCAGGAGCTTCTGTAGA-3‘) and mamm800r (5‘-GACATAGCTGGAGGTTTTATGT-3‘) to produce a 121-bp amplicon of the CO1 gene. The specific PCR conditions are described in the supplementary material. Mammal metabarcoding PCRs were performed on 21 DNA extracts of core LK-001 and 27 samples of core LK-007, with eight independent replicates, each. Each batch of PCRs included one non-template control. Established metabarcoding primers were used (Giguet-Covex et al., 2014), and human blocking primers (Garcés-Pastor et al., 2021) were included. PCR conditions are described in the supplementary material. Sequencing was performed on an Illumina NovaSeq platform, with 2 x 150 reads. The raw data were processed as described in the supplement. We used the sample LK-001_66.5 of core LK-001 which had produced one of the highest Mammuthus read counts after enrichment. The specific target amplified by the primer pair were used to design a probe (5‘-GGATACTCCTGCAAGGTGAAGTG-3‘). With this probe, ddPCR was performed with 24 extracts of core LK-001 and with 30 extracts of core LK-007, with three replicates, each (supplement).

Supporting information

eLife Assessment

This work presents convincing evidence for the presence of wooly mammoth/rhinoceros ancient environmental DNA (aeDNA) far from the time likely to host living individuals: what is effectively a genetic version of a geological inclusion. These are important findings that will have ramifications for the interpretation and conclusions extracted from aeDNA more generally.

Reviewer #1 (Public Review):

This manuscript presents an important study that contributes to our understanding of the reliability of ancient environmental DNA (aeDNA) extracted from sediment cores. The authors address the potential biases and challenges associated with using aeDNA to infer past ecosystems, specifically focusing on the case of mammoths and woolly rhinoceroses in the Yamal peninsula, West Siberia.

The introduction provides an overview of the significance of sedimentary deposits as archives of past ecosystem changes and illustrates the remarkable insights gained from previous studies using aeDNA, highlighting its potential for reconstructing paleoecology, phylogeography, and understanding extirpation and extinction events of keystone taxa. The authors then report the detection of DNA and near complete mitochondrial genomes of multiple mammoth and woolly rhinoceros individuals in the sampled sediment cores (which are dated to the last few centuries). The authors then employed additional methods to confirm the presence of ancient DNA from mammoths in these sediment cores. Conventional PCR and Sanger sequencing of a mammoth COI fragment confirmed the amplification of mammoth DNA. Mammal metabarcoding and droplet digital PCR (ddPCR) further supported the detection of mammoth DNA in both cores.

The hybridisation enrichment experiment results showed high read counts assigned to Mammuthus, ranging from 2,852 to 72,919 reads per library in core LK-001. Negative controls did not produce any reads assigned to mammals, indicating the absence of contamination. The study also revealed the presence of woolly rhinoceros sequences in the sediment cores, with 12 out of 23 libraries producing more than 100 reads assigned to woolly rhinoceros. The total number of woolly rhinoceros reads was 2,737, and the cumulative mitogenome coverage reached 44%.

The authors carefully addressed the incongruity between the temporal occurrence of these extinct species and the presence of their DNA in recent sediments. They proposed several mechanisms that could explain the recovery of Pleistocene megafaunal DNA in the sediment cores. The minor amount of ancient DNA post mortem damage observed in the mammoth sequences indicates exceptional preservation, consistent with an origin from (recent) permafrost. The dynamics of permafrost thawing and redeposition in the study area provide a plausible explanation for the presence of ancient DNA in the sediments. The authors discuss potential mechanisms for the redistribution of Late Pleistocene material in the sediments, including thermo-denudation processes, methane emissions from degrading permafrost, and the formation of taliks and methane seepage. These processes can disturb the stratigraphy of lake sediments and potentially mix ancient material within the modern sediments. I believe the conclusions are supported by the data and the manuscript is well-written and clear to follow for the reader.

Reviewer #2 (Public Review):

Summary:

The authors report the successful retrieval of mitogenomes from extinct Pleistocene megafauna (woolly Mammoth and woolly rhino) from recent sediment cores from two close Siberian lakes. The cores are too recent to represent real time points of these two extinct species (known to have been extinct for several thousands of years) and therefore, the most plausible interpretation is that permafrost thawing and similar physical processes in the lakes have made surface old ancient DNA, maybe from nearby, deep-buried carcasses.

Strengths:

The pattern of postmortem damage at the end of the Mammoth DNA reads as well as the length distribution (reported in Figure 1) is expected for authentic ancient DNA extracts (besides the phylogenetic evidence). These results pose a question, in my view, on the general reliability of sedimentary DNA in similar contexts, especially in the absence of direct radiocarbon dating of associated remains and in the absence of an understanding of the local geo-physical dynamics. At the same time, the evidence reported here suggests that, at least in Siberian lakes, the sediments can preserve a rich ancient DNA record that it is worth surveying.

Weaknesses:

Although admittedly the work can represent two cases of environments with singular thermal conditions and geodynamics, it opens also the possibility of studying more lake sediments for trying to understand if these findings can be generalized.

Reviewer #3 (Public Review):

Summary:

In this study, the researchers used ancient environmental DNA (aeDNA) retrieved from sediment cores, from two lakes in the Arctic, on the Yamal peninsula, in Siberia. The dating of one of the cores, showed that the sediment layers were very recent (ranging between the years 2019 - 1895). From this core they sequenced 23 libraries which were enriched for mammal mitochondrial genomes. They found a high proportion of two species that have been extinct for thousands of years, the mammoth and the woolly rhinoceros. The highest proportion of mammoth reads were found in very young layer (~81 years old) and as this initial finding does not match the temporal occurrence of the species, they confirmed the identification with several other methods. Additionally, they applied a different dating method on some samples and found that the aging of the samples was not completely congruent. The authors suggest the that the presence of these two Pleistocene megafauna in such recent sediment layers is a consequence of physical processes, specific to the study site, and that the high quality of the aeDNA recovered is a result of permafrost preservation.

Strengths:

The strengths of the study are in the rigorous confirmation of the identification of the taxa with four different PCR and sequencing techniques being used, the initial enrichment panel, and then subsequent metabarcoding PCRs, and taxa specific PCR for COI and cytB. Along with the ancient DNA protocol applied, this is therefore very convincing that the DNA detected in the samples is indeed from the Pleistocene mammals. Additionally, two methods were used to age the sediment cores, and although the depth of the samples tested do not overlap, they give reasonable ages (apart from the anomalous sample) and all together these are robust results.

Weaknesses:

The paper could benefit from clearer aims in the introductions because as it stands the initial aim states that the authors are looking for Arctic mammal abundances through time. However, there are no results relating to general arctic mammal biodiversity presented, which leaves the reader wondering. Perhaps the focus of the study is more on identifying and dating the Pleistocene megafauna. Additionally, it is presented as an analysis on the two taxa, but it feels like the woolly rhinoceros does not receive the same treatment as the mammoth, as there are no additional molecular results, confirmation or figures relating to DNA from this taxa.

Overall the results support that there has been some movement of DNA throughout the sediment core which may impact the dating of the last occurrence of particular extinct taxa. As highlighted, though the geological processes by which this may have arisen are specific to this particular lake and may not be broadly relevant, therefore highlighting that knowledge of each system is important to understanding DNA distribution.

Author Response

The following is the authors’ response to the original reviews.

The reviewers make some suggestions aimed towards increasing the clarity of the manuscript, and I suggest that the authors examine those carefully. In particular, the figure is difficult to read and could contain additional information to help the reader's interpretation. For example, Reviewer 1 suggests including sample age estimates alongside depth, while Reviewer 3 also notes that there is missing information in the figure. Apart from the figure, Reviewer 1 suggests two additional analysis to help explain the amount of mammoth DNA recovered, which they observe is much higher than previous similar investigations. This would seem to be an important issue to address, given the surprising nature of the findings. In addition to this larger issue, the Reviewer makes a few important suggestions for supplementary material that may be needed to support the authors' statements.

Some additional recommended edits -- in particular to the text and included references to related studies -- are suggested by Reviewers 2 and 3, and both commented on the lack of a publicly-available data repository. The authors may also wish to comment on or revisit their differential treatment of wooly mammoth vs. wooly rhinoceros samples, though I suspect this has more to do with low read numbers for the rhinos.

Thank you very much for the positive assessment of our manuscript and clear suggestions for revision. We address these points below.

Reviewer #1 (Recommendations For The Authors):

I have a few suggestions that might further improve the manuscript:

It is difficult for the reader to follow which core slices exactly have been sampled and sequenced. The authors mention 23 samples were taken from core LK-001 and 16 samples from core LK-007. From the text it remains unclear to me what the exact age of each of these samples is. Figure 1 shows the depth at which the LK-001 core was sampled, maybe sample age estimates could be included here.

Thanks for pointing this out. We have added approximate ages to Figure 1, added the depth range to the text (“from 1.5 to 80 cm”; l. 73-74, caption Figure 1), and reworked the table of the sampling depths in the supplement.

Line 84-87. The authors mention the retrieval of DNA from several expected Arctic taxa, however no further data regarding these findings is given in the manuscript. It would be useful to report the same numbers for these species as the ones given for the Mammuthus and woolly rhinoceros, which would allow for a comparison of the relative abundance of the DNA between these species. Are the expected Arctic species for instance at much higher (DNA) abundance in the samples? It would also be interesting to know if the authors discovered DNA from extant species that are unlikely to have occurred in the geographic region. A (supplementary)table listing the number of mapped reads to each of the respective mitogenomes for each sequence library would be useful for the reader.

We added a supplementary table (S8) indicating the numbers of reads assigned to mammals.

Line 90: I am somewhat amazed by the amount of mammoth DNA the authors recovered from these cores. A total depth of over 400X of the mitogenome is quite extraordinary and I am not aware of any ancient sediment study to date that has retrieved a similar amount of data. For instance, the Wang et al. 2021 paper, which the authors cite, sequenced over 400 samples and did not find any mammoth DNA in 70% of those. For the 30% of samples showing signs of mammoth DNA they retrieved on average 530 sequence reads. In this study the authors find on average ~20.000 reads, in 22 out of the 23 sequence libraries. This makes me wonder if the way the mapping was performed has been too lenient, resulting in possible spurious mappings? To really confirm the authenticity of the mammoth (and woolly rhino data) I would suggest two additional analysis:

1. Mapping all the sequence libraries to a reference consisting of the complete Asian-elephant genome (for instance https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_024166365.1/), the complete human genome (+mitogenome) and the Asian elephant mitogenome. This could possibly reduce spurious mappings as conserved regions between the genomes are filtered out and could also reduce the possible mapping of NUMTS. If the authors could show that after such a mapping approach a significant number of reads are still assigned to the Asian elephant part (including the mitogenome) of the reference, the reported findings would be strengthened.

1. I also suggest to construct a mitochondrial haplotype network from the obtained DNA, while also including previously published Asian and African elephants as well as previously published mammoth mitogenomes. If the obtained haplotypes indeed show that they cluster within the known haplotype diversity of mammoth, that would be strong support for the authenticity of the data

The same analysis could be considered for the woolly rhino data, although the lower read numbers might make this analysis challenging.

We agree that the amount of mammoth DNA is surprising, which is why we opted for further laboratory experiments for confirmation of the hybridization capture results of the first core, i.e., 1) DNA extraction from a second core of a different lake, 2) a quantitative PCR approach (ddPCR), and 3) metabarcoding. Our results of the highly specific ddPCR and metabarcoding assays confirmed considerable amounts of mammoth DNA in two sediment cores of different lakes, thus we have no doubts regarding the authenticity of the data. Considering the large amount of mammoth DNA, the high number of reads, and particularly the high mitogenome coverage, we argue that the effect of some spurious mapping is negligible and does not affect the main outcome and conclusions of our study. Although we agree that a haplotype network would be interesting, such analyses would stretch beyond the focus of this publication.

Line 91: The authors mention negative controls (extraction and library blanks) did not produce any reads assigned to mammals. This is quite remarkable, as in my experience low levels of (human)contamination are almost always present in the blanks. Could the authors comment on why they think the blanks did not show any signal of mammalian DNA?

The hybridization capture enrichment and the filtration and mapping procedures likely eliminated human contamination. Also, the data were mapped against Arctic mammal mitogenomes, which did not include human reference sequences. However, six of the sediment samples contained human sequences (now shown in supplementary table S8), albeit at low read counts (mean = 65)

Line 97: "mapping suggested that the sequences throughout the core originated from multiple individuals" The authors do not provide any supporting data showing this. I think that an analysis (for instance based on allele frequencies) has to be included in manuscript to support this claim.

We agree that his claim was not sufficiently supported. We performed further analyses including genomic data of previously retrieved mammoth remains and assigned our data to these haplogroups; the results were added to the main text and are shown as a figure (Fig. 2).

Line 98: "Signatures of post-mortem DNA decay were comparably minor."

Do the authors know if the used hybridisation enrichment method can distort the measurement of post-mortem damage? Are for instance reads with C-T substitutions less likely to be captured by the baits?

To our knowledge, there is no study suggesting that damaged sites are less likely to be captured. In general, the hybridization capture procedure is not overly specific, and studies report that DNA is readily and preferentially captured as long as the difference between baits and DNA is not above 10%.

Line 100: "The proportions of bases did not suggest a substantial deviation from those in the reference genomes or in the closest extant relative of Mammuthus, the Asian elephant (Elephas maximus)."

It is not clear to me what the authors mean by this. Could the authors explain how this was measured and what their interpretation of this result is?

We realize that the sentence was unclear. We meant that the nucleotide composition was similar to that of the reference genomes or the closest extant relative. However, as we do not consider this important for the argument, we have removed this sentence from the manuscript.

Given the high number of recovered mammoth reads in the samples, it would be interesting to know how much mammoth reads are present in the sample before enrichment capture with the baits. Shotgun sequencing the raw extract of one of the samples with the highest number of mammoth reads might allow for a rough estimate of mammoth DNA abundance compared to the other extant species (e.g. reindeer, Arctic lemming and hare) found in the sample(s). This could give further clarification about the extent of stratigraphy disturbance and its overall effect on the DNA based community reconstruction. However, this is just a suggested additional analysis and not something I believe crucial for supporting the overall findings in this manuscript.

We fully agree that this would be a highly interesting and informative additional analysis to perform. It was, however, not possible to perform this additional analyses in the course of the current experiments.

Finally, I could not find a public link to the (sequence)data produced in this study. I strongly encourage the authors to make their data publicly available.

Thank you for pointing this out. We have added a Data Availability paragraph, including the respective reference.

Reviewer #2 (Recommendations For The Authors):

In the Discussion it is mentioned that the reasons for Mammoth extinction are not entirely clear but are largely attributed to sudden climate warming (and add some relevant citations). However, there is also abundant literature that suggest humans also played a role in their extinction (for instance, a recent one, Damien et al. (2022) at Ecology Letters 25: 127-137).

We agree with the reviewer and have added some the recent citation highlighting the possible influence of humans.

One possibility to add further interest to this paper would be to conduct a phylogenetic tree with the Mammoth mitogenome(s) retrieved and a reference dataset; it could be interesting to know where do they fall in the phylogeny -already abundant with tens of individuals- and maybe it could be even possible to roughly estimate their date. There are some papers that report many Mammoth mitogenomes, including of course some from Siberia; for instance Chang et al. (2017) at Sci Reports and also Fellow Yates et al. (2017) also at Sci Reports (the latter mainly from Central Europe).

We are well aware of the amount of mt genomes available for mammoth, and such an analyses would be an interesting addition, potentially also offering the possibility to date the DNA. However, the analyses was hampered and would be less secure for this dataset, as our sequences display quite some variation among each other, suggesting that we have a mix of multiple mt genomes, which we cannot readily distinguish. We thus refrain from this, also because we instead provide multiple lines of evidence for the existence of the mammoth DNA in the surface sediment core (metabarcoding, ddPCR).

Minor points:

-Correct wooly to woolly

Revised.

-In the sampling description it is not totally clear if the samples were taken at 1 cm each (it is mentioned that core LK-001 is sliced in the field at 1-cm steps for radiometric dating and later it is explained that 23 samples were analyzed from this core, but it is unclear if they represent 23 cm of core)

-Maybe the authors could briefly define some terms such as "talik"

Revised.

Reviewer #3 (Recommendations For The Authors):

Maybe I missed this but I could not find a data availability statement or the location of the repository

We have added a Data Availability paragraph, including the respective reference.

It would be good to see some additional analysis on the distribution of the woolly rhinoceros DNA through the sediment core - like the figure for the mammoth i.e read numbers vs depth.

We have added to the supplements a table showing the numbers of assigned mammal reads over the core depths (Table S8). However, as rhinoceros reads are considerable rarer in our results, we did not produce a figure.

Would it be possible to be more explicit about the multiple mammoth individuals, could you calculate a minimum number or haplotypes for example.

We agree that his claim was not sufficiently supported and added results from additional analyses (incl. Fig. 2). Please see our response above.

Based on the aim stated in the introduction, the analysis of the Arctic biodiversity of this area is missing, it would be nice to see these result added or maybe the focus needs to be changed for clarity.

We now explicitly state that this objective pertains to a different study, which is currently still in preparation for publication.

The single main figure needs a bit more consideration. For example in panel A - there was no information on the transformation performed or what the general trend line refers to. Do the results in panel B refer to all 22 libraries? What is the x-axis in Panel C and what do the coloured lines refer to? Additionally, I think the figure needs to be in higher resolution with increased text size on all axes.

We revised the figure and the caption for clarity and readability.

Finally this might be an accidental typo - but when referring to the sample aged at around 8,677 years in text it states this the 36.5 cm sample (line 130 and 192), but the supplementary says this is the 51cm sample (Table S6). This would maybe impact potential conclusions. Would you be able to clarify this.

Thank you for noting this error, we revised it.

Version 2

Abstract

Abstract

Ancient environmental DNA (aeDNA) from lake sediments has yielded remarkable insights for the reconstruction of past ecosystems, including suggestions of late survival of extinct species. However, translocation and lateral inflow of DNA in sediments can potentially distort the stratigraphic signal of the DNA. Using three different approaches on two short lake sediment cores of the Yamal peninsula, West Siberia, with ages spanning only the past hundreds of years, we detect DNA and identified mitochondrial genomes of multiple mammoth and woolly rhinoceros individuals—both species that have been extinct for thousands of years on the mainland. The occurrence of clearly identifiable aeDNA of extinct Pleistocene megafauna (e.g., > 400K reads in one core) throughout these two short subsurface cores, along with specificities of sedimentology and dating, confirm that processes acting on regional scales, such as extensive permafrost thawing, can influence the aeDNA record and should be accounted for in aeDNA paleoecology.

Introduction

Sedimentary deposits constitute highly valuable archives of past ecosystem changes as they contain dateable layers with organismic remains including ancient DNA (aDNA). Such remains are typically assumed to represent the ecosystem of the time around which the respective stratum was deposited. aDNA from sediments has yielded remarkable insights regarding paleoecology, phylogeography, and extirpation and extinction events of keystone taxa such as mammoths (Haile et al., 2009; Boessenkool et al., 2012; Graham et al., 2016; Murchie, Monteath, et al., 2021). Based on such ancient environmental DNA (aeDNA), a recent study proposed that the woolly mammoth (Mammuthus primigenius) may have survived in Eurasia for much longer than previously assumed, as the authors retrieved mammoth DNA sequences in sediment layers that were approximately 4.6–7 thousand years (kyr) younger than the most recent mammoth fossils (Wang et al., 2021); however, in response to this interpretation, Miller and Simpson (2022) opined that these results may be more likely due to taphonomic processes leading to release of aeDNA from the remains of long-dead organisms from permafrost, where it is well preserved.

Conclusions derived from aeDNA isolated from sediment cores rely on the stratified nature of the remains in question and dating of the respective layer by radiometric methods. However, in theory, various physical and geochemical processes such as translocation of DNA through sediment strata (Haile et al., 2007), re-deposition of older sediment carrying DNA of extinct organisms (Arnold et al., 2011), and preservation bias (Boere et al., 2011) can distort the biological signal of aeDNA and thus bias the accuracy of allocation of taxa to specific time periods (Armbrecht et al., 2019). For lake sediments, deposited under aquatic conditions, studies have suggested that leaching is not a concern (Parducci et al., 2017), but it has been observed in soils and cave sediments (Haile et al., 2007). The question of obtaining last appearance dates of extinct taxa using aeDNA in dated sediment layers (Haile et al., 2009) is under discussion, but the surprisingly young records published so far still date to multiple thousands of years before present and thus lie within a timeframe of possible late survival.

Results and Discussion

In 2019, we retrieved short subsurface sediment cores from two Arctic thermokarst lakes (LK-001 and LK-007, located approximately 5 km apart, over massive permafrost; Table 1; Dvornikov et al., 2016) on the Yamal peninsula, Siberia, to extract DNA and assess changes in mammal abundances in the Arctic over the past decades and centuries. From lake LK-001, we collected a secondary core which was sliced in the field at 1-cm steps for Pb210 radiometric dating, which indicated that the sediments at the top of this core were deposited recently, and that the core spanned the past few centuries (Table S7). The cores for DNA extraction were closed in the field immediately after retrieval and were then transported to the dedicated aDNA laboratories of the University of Konstanz, Germany. In this lab facility, established in 2020, no other samples from the Arctic or from any large mammals had been processed previously. From core LK-001, we isolated DNA and produced genomic double-stranded libraries from 23 samples, from 1.5 to 80 cm core depth (Supplement section 1), according to standard procedures. The core was opened and all subsequent steps until index PCR setup were carried out under customary aDNA laboratory conditions. In particular, the core opening facilities and the lab are located in buildings separated from the downstream molecular genetic analyses, the ventilation of the aDNA lab is based on a HEPA filter system and positive air pressure, and the lab is subjected to nightly UV radiation. Work in the lab is conducted under strict aDNA precautions, adhering to established aDNA protocols (Fulton and Shapiro, 2019). We enriched the libraries for mammalian DNA using a custom RNA bait panel produced from complete mitogenome sequences of 17 mammal species that currently or previously occurred in the Arctic (adapted from Murchie, Kuch, et al., 2021). The enriched libraries were sequenced, and we mapped the sequences against a database of 73 mammal mitogenomes, followed by BLASTn alignment against the complete NCBI nt database. We thus retrieved mitogenomic sequences of mammals that were expected during the age range covered by the core (Table S9), e.g., reindeer (Rangifer tarandus), Arctic lemming (Dicrostonyx torquatus), and hare (Lepus); however, throughout the entire core, there were abundant sequences of two species that have been extinct for several thousand years, i.e., mammoth (Mammuthus primigenius) and woolly rhinoceros (Coelodonta antiquitatis). Twenty-two of the 23 LK-001 libraries produced > 1,000 reads, each, assigned to Mammuthus, with read counts ranging from 2,852 to 72,919 (mean 21,140 ± 17,296). Negative controls (extraction and library blanks) did not produce any reads assigned to mammals. In the sample with the highest M. primigenius read counts (31.5 cm depth, dated to 81 years), the coverage of the reference mitogenome (NCBI accession NC_007596.2) was 95.3%, (434 (± 213)-fold). Across all samples, 465,080 reads assigned to mammoth were produced, with 98.3% coverage (2,762-± 1,176-fold). Read lengths ranged from 28 to 289 bp (mean 100 ± 44 bp; Fig. 1). The number of woolly mammoth reads decreased from lower samples towards the top of the core (Fig. 1). Signatures of post-mortem DNA decay were comparably minor (Fig. 1), with reference to an M. primigenius genome downloaded from NCBI (accession NC_007596.2), and mapping suggested that the sequences throughout the core originated from multiple individuals. Further analyses of the three libraries with the most mammoth reads using mixemt (Vohr et al., 2017) identified a number of mitochondrial haplogroups in the sequences from the core, suggesting that they originated from a multiple individuals (Fig. 2.). The haplogroups identified were known to occupy the region, and it seems likely the sequences reflect a history of mammoth occupation at the core site. Twelve of the 23 libraries produced > 100 reads, each, assigned to woolly rhinoceros, with a total of 2,737 reads and a cumulative coverage of 44% (assembled to NC_012681.1 2). Further analyses are mostly focused on the mammoth sequences as these occurred in substantially higher numbers.

aDNA of Mammuthus in recent lake sediments. (A) Read counts assigned to Mammuthus (square-root-transformed proportion of the respective number of raw reads per library) after hybridization capture enrichment of aeDNA of core LK-001 (shown are results of 22 libraries; one library was excluded as it did not produce any reads assigned to mammals); square-root transformation of percentage. Indicated are sample depths (in cm; 1.5 to 80 cm) and approximate ages as per 210 Pb chronology (Table S7; to a maximum depth of 39.5 cm). The solid line indicates the general trend. Across the 22 libraries: (B) Fragment length distribution and (C) damage patterns (red indicates C-to-T transitions, blue G-to-A transitions. the Y-axis indicates the percentage of positions with a nucleotide change, the X-axis indicates the position along the fragment).

>. Locations of the sediment cores of the present study (Yamal peninsula, Siberia) and previously retrieved mammoth remains and their haplo(sub)groups (Table S6). The bar chart indicates a maximum-likelihood estimate of the haplogroup proportions derived from the reads from the three sediment core libraries with the most mammoth reads.

As these results were entirely incongruent with the temporal occurrence of these two species, we employed several methods to confirm their validity, i.e., conventional PCR and Sanger sequencing of a mammoth cytochrome c oxidase subunit I (COI) fragment, mammal metabarcoding, and droplet digital PCR (ddPCR) of a mammoth cytochrome b fragment. We performed these analyses on core LK-001 and, to evaluate whether this is a locally isolated phenomenon, on a second short core, LK-007, from a nearby lake, which had been opened and processed under the same conditions indicated above to prevent contamination. Additionally, we screened the core LK-001 for plant macrofossils, of which three were sent for 14C AMS dating.

Conventional PCR and Sanger sequencing confirmed amplification of a COI fragment of M. primigenius. Mammal metabarcoding produced mammoth sequences (74 bp) in 13 samples of core LK-001 and 9 samples of core LK-007. In the LK-001 core, the highest M. primigenius read count occurred at 31.5 cm (2,992 reads), and in the LK-007 core at 26 cm M. primigenius (3,580 reads). ddPCR produced M. primigenius sequences in 14 samples of each core. Metabarcoding and ddPCR patterns across the cores were similar, although not completely congruent, as ddPCR appeared to be more sensitive (Fig. S1). The dates retrieved by radiocarbon dating were not congruent with the initial age inference suggested by Pb210. While the lowest and topmost sample (with ages of 1,547 ± 228 and 339 ± 79 uncal. yrs BP respectively) suggest relatively young ages and agree in their temporal succession, the middle sample, at 51 cm, yielded a radiocarbon age of 8,677 ± 132 yrs.

The mammoth was abundant throughout most of Eurasia during the Pleistocene, but populations declined at the end of the Pleistocene, with the species going extinct in the mid-Holocene (Nogués-Bravo et al., 2008). The youngest fossils from mainland Siberia have been dated to 9,650 years (Stuart et al., 2002). The woolly rhinoceros was a cold-adapted megaherbivore, which was abundant from western Europe to north-east Siberia during the Middle to Late Pleistocene (Kahlke and Lacombat, 2008). This species was predominantly grazing, probably resorting to browsing only due to seasonal vegetation restrictions (Kahlke and Lacombat, 2008; Rey-Iglesia et al., 2021; Stefaniak et al., 2021). The reasons underlying the extinction of this species at ca 14 ka BP are not entirely clear, but it is largely attributed to sudden climate warming and subsequent habitat unsuitability due to vegetation changes (Stuart and Lister, 2012; Lord et al., 2020; Wang et al., 2021), likely coupled with human influence (Fordham et al. 2022), in the Bølling-Allerød interstadial warming.

The present data, which implies frequent and abundant Pleistocene megafaunal DNA throughout a sediment core deposited in the lake over the past centuries suggests that physical processes, rather than presence of live organisms, are responsible for the recovery of this DNA. While not in itself fully conclusive, our data suggests the source of the DNA of Pleistocene mammals from older permafrost deposits, either from a carcass or from sedimentary materials carrying the DNA. The numbers of mammoth sequences increased with depth downcore, with comparably low abundances over the most recent 11 cm of the core (aged < 30 years), pointing to a decrease of input of this DNA through time. The apparently rather limited extent of aDNA damage in the mammoth sequences suggests that the source of this DNA has been preserved exceptionally well, which also suggests an origin from permafrost, and the specific dynamics of thawing and re-deposition of material in the study area offer an explanation.

Here, the active thermokarst likely began during the climatic optimum of the Holocene (9000–3000 BC; Savvichev et al., 2021). The LK-001 and LK-007 lake basins are interbedded mainly into the IVth marine plain, formed in the Kazantsevskaya, Marine Isotopic Stage 5. The permafrost of these lake catchments was formed no earlier than 70—60 kyr BP after the Kazantsevskaya transgression in more sub-aerial conditions and with mean annual temperatures 6 to 7 °C lower than modern temperatures (Baulin et al., 1981). As this area was under coastal-marine conditions for a long period, these lake basins may be paleo-marine remnants, or they were formed later as a result of thermokarst over the segregated or tabular (massive) ground ice during the Holocene climatic optimum (Kachurin, 1961). The area is also subject to abrupt permafrost thaw (thermo-denudation), resulting, for example, in the formation of retrogressive-thaw slumps (thermocirques) and the transport of a large amounts of thawed terrestrial material into the lake water (Dvornikov et al., 2018). Such abrupt permafrost thaw processes normally appear adjacent to lakes and can form specific geomorphological elements,. i.e., thermo-terraces (Kizyakov et al., 2006) within lake catchments and lakes; they are normally polycyclic processes appearing due to the extension of a seasonally thawed layer (active layer) up to the top of massive ice and more favorable thermal conditions within the existing forms (Leibman and Kizyakov, 2007; Kokelj et al., 2009). Traces of past thermo-denudation can be observed within both lake catchments. In catchments of five neighboring lakes, large retrogressive-thaw slumps appeared in recent years (2012-2013) accompanied by the thaw and lateral transport of modern and Late Pleistocene deposits into lakes. Additionally or alternatively, thermo-erosion of upper geomorphological levels and transport via stream networks could transport ancient material into the modern lacustrine sediments. However, the two studied lakes are headwater lakes (with outlet, no apparent inlet) and this option can only be considered in terms of small thermo-erosional valleys within the catchments.

An alternative mechanism for the redistribution of Late Pleistocene material in the sediments is related to subcap methane emission (bubbling) from degrading permafrost beneath the lake bottom. In-lake bubbling can be observed in a circum-Arctic scale: in North-East Siberia, Alaska and Canada. This is common especially in lakes with a depth exceeding two meters, which do not freeze entirely up to the bottom in winter, leading to the formation of a talik (a layer of year-round unfrozen ground that lies in permafrost area). The expansion of the talik may further trigger subcap methane emission, which can reach 40-70 kg yr-1 of pure (94%–100%) methane in neighboring lakes (Kazantsev et al., 2020). The constant methane seepage doesn’t allow ice to be formed in winter (whereas the normal winter ice thickness is approximately 1.5 m) and can potentially disturb the stratigraphy of lake sediments. Additionally, dramatic emissions of methane can form craters in terrestrial and lacustrine environments (Dvornikov et al., 2019). In this case Late Pleistocene sediments will well be re-distributed within the water-body and the entire stratigraphy will be mixed. In the case reported here, the simultaneous finding of a Pb210 chronology indicating recent sediment deposition and of plant macrofossils that dated to > 8000 yrs BP in a sample from 36.5 cm, suggest lateral input of ancient material, including the mammalian DNA, putatively related to permafrost thawing processes. Numerous studies on aDNA discussed possible leaching through sedimentary strata of the DNA itself, yet it was typically considered not an issue as most of these studies were conducted under stable permafrost or similar conditions (e.g., Hansen et al., 2006; Haile et al., 2007, but see Andersen et al. 2012). Permafrost thawing and re-deposition of material adds a new dimension to this problem of temporal interpretation. The fact that we retrieved mammoth sequences from both cores of two lakes located approximately 5 km apart suggests that this is not an isolated phenomenon but occurs on a regional or even larger scale. Given the wide spread of abrupt permafrost thaw processes in the Arctic plains (West Siberia, Taimyr, Chukotka, Alaska, Canadian Arctic; Kizyakov et al., 2006 and references therein), the phenomena of disturbed stratigraphy of lacustrine sediments can potentially be observed at a pan-Arctic scale. While this indicates that temporal interpretation of sedimentary aeDNA records should be exercised with caution, our study also demonstrates that a careful evaluation of available information on the site and ecosystem in conjunction with the use of independent dating techniques can uncover incongruencies. This is more difficult in older time periods, where artefactual stratigraphies caused by equivalent processes acting for a limited time will not be detected as easily as in our case of long extinct species. The same applies to extant taxa or those which have undergone extinction or extirpation more recently, the presence of which cannot as easily be excluded as in the current example. However, we suggest that the inclusion of robust dating techniques and knowledge of local geophysical processes can provide good arguments to evaluate the reliability of aeDNA records.

Materials and Methods

Field sites, DNA isolation and hybridization capture enrichment

In 2019, sediment cores (6 cm diameter; Table 1) were retrieved from two lakes (LK-001 and LK-007, respectively, Table 1) on the Yamal peninsula, Siberia, using a UWITEC piston corer (UWITEC, Mondsee, Austria). The lakes were located approximately 5 km apart. The cores were transported to the aDNA laboratories of the University of Konstanz, Germany; from lake LK-001, a secondary core was taken which was sliced in the field at 1-cm steps for radiometric dating from 0 to 39.5 cm depth, performed at the Environmental Radioactivity Research Centre, University of Liverpool (Supplement section 9). Additional 14C dating of three specimens of plant remains, extracted at 36.5 cm, 51 cm and 74 cm, was performed (Supplement section 9).

Sedimentary DNA was isolated from 23 samples of core LK-001 and from 16 samples of core LK-007 using commercially available kits with modified protocols (Supplement section 1). The extracts of core LK-001 were subjected to library preparation for capture enrichment. Enrichment probes were designed from mitogenomes of 17 herbivorous mammal species that currently or previously occurred in the Arctic (Table S2) and few lichen sequences (Table S3). Genomic libraries were produced according to Li et al., 2013 with some modifications (Seeber et al., 2019; Seeber et al. 2023; Supplement section 2). Filtered reads were mapped to mammalian mitogenomes, followed by BLASTn alignment against the complete NCBI nucleotide database and subsequent metagenomic analyses using MEGAN (Huson et al., 2016). Reads assigned to Mammuthus were mapped to a complete M. primigenius reference mitogenome (NCBI accession NC_007596.2); reads assigned to Coelodonta antiquitatis were mapped to the NCBI reference mitogenome NC_012681.1 2. The reads mapped to mammoth from the top three libraries were assigned to haplogroup using mixemt (https://github.com/svohr/mixemt) with a custom-made representative panel of 15 mammoth mitogenomes (Fig. 2; Table S6).

Conventional PCR, mammal metabarcoding, and ddPCR

Based on the enriched fragments with the highest coverage, PCR primers specific to M. primigenius were designed using Geneious Prime 2022.1.1 (Kearse et al., 2012), i.e., mamm801 (5’-CCCATGCAGGAGCTTCTGTAGA-3’) and mamm800r (5’-GACATAGCTGGAGGTTTTATGT-3’) to produce a 121-bp amplicon of the CO1 gene. The specific PCR conditions are described in Supplement section 6. Mammal metabarcoding PCRs were performed on 21 DNA extracts of core LK-001 and 27 samples of core LK-007, with eight independent replicates, each. Each batch of PCRs included one non-template control. Established metabarcoding primers were used (Giguet-Covex et al., 2014), and human blocking primers (Garcés-Pastor et al., 2021) were included. PCR conditions are described in the supplementary material. Sequencing was performed on an Illumina NovaSeq platform, with 2 x 150 reads. The raw data were processed as described in the supplement. We used the sample LK-001_66.5 of core LK-001 which had produced one of the highest Mammuthus read counts after enrichment. The specific target amplified by the primer pair were used to design a probe (5’-GGATACTCCTGCAAGGTGAAGTG-3’). With this probe, ddPCR was performed with 24 extracts of core LK-001 and with 30 extracts of core LK-007, with three replicates, each (Supplement section 8).

Supporting information